---
title: "maldiscrim-demo"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{maldiscrim-demo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width  = 10,
  fig.height = 6,
  out.width  = "100%",
  fig.align  = "center"
)
```

## Introduction and Scientific Context

In microbiology, discriminating bacterial samples at a fine taxonomic resolution—specifically at the **strain level**—remains a major challenge. While highly accurate molecular tools exist, their exorbitant costs and complex workflows often hinder routine laboratory use and large-scale industrial screens. 

Matrix-Assisted Laser Desorption/Ionization Time-of-Flight (**MALDI-TOF**) mass spectrometry has emerged as an excellent alternative due to its cost-effectiveness, speed, and widespread availability in routine laboratories. However, standard software and traditional workflows typically restrict identification to the **genus or species level**. This limitation stems from a heavy reliance on "peak-picking" algorithms, which isolate only the most intense spectral peaks. While peak-picking captures the primary signal, it discards a massive amount of low-intensity background data and baseline structures. This omitted information turns out to be crucial for capturing the subtle biological signatures that differentiate one strain from another within the same species.

The `maldiscrim` package bridges this gap by shifting the paradigm from peak analysis to **Functional Data Analysis (FDA)**. By treating the entire MALDI-TOF mass spectrum as a continuous functional curve in its **completeness**, this package extracts the hidden information located outside the major peaks. 

Combined with robust functional decomposition (B-splines or Wavelets) and discriminant modeling (fPLS-DA), `maldiscrim` provides microbiologists with an accessible, affordable, and highly accurate statistical framework to achieve strain-level distinction without the financial burden of expensive alternative technologies.

This vignette demonstrates the complete analysis workflow, from raw spectrum preprocessing to the visualization of discriminant latent spaces.

---

### 1. Data Loading and Preprocessing (`ProcessMALDI`)

The `ProcessMALDI` function imports raw spectra files from a target directory. It encapsulates a rigorous pipeline powered by `MALDIquant`: variance stabilization, Savitzky-Golay smoothing, baseline removal (`SNIP`), and Total Ion Current (`TIC`) intensity normalization.

Technical replicates are aligned and averaged into a single **biological replicate** (Mean Spectrum Profile, or MSP) by combining the sample's full name and its acquisition month. The output is a clean matrix of continuous profiles.

```{r load-package}
library(maldiscrim)
```

```{r process-maldi, eval = FALSE}
# Example of importing and preprocessing raw spectra files
# Each row represents a biological replicate (MSP) interpolated across 20,664 m/z points
data_path    <- "path/to/your/strains_directory"
raw_processed <- ProcessMALDI(dataFile = data_path, methodBaseline = "SNIP", halfWindowSize = 10)
```

For the remainder of this vignette, we will use the package's built-in reference dataset.

```{r load-data}
data("spectra100")

# 100 biological replicates x 20,664 m/z channels
dim(spectra100)
```

---

### 2. Training the Classification Model (fitFPLS_DA)

The fitFPLS_DA function fits a supervised classifier tailored for functional data. It projects the continuous signal onto a functional basis (B-splines or Wavelets) and extracts expansion coefficients to train a multivariate Partial Least Squares (PLS) regression.

The optimal number of PLS components is automatically selected via cross-validation using the One-Sigma rule. Finally, a Linear Discriminant Analysis model (MASS::lda) is fitted on the PLS scores, which are filtered beforehand to remove columns with zero variance.

```{r fit-model, message = FALSE, warning = FALSE}
# Fitting the FPLS-DA model using a B-splines functional decomposition
fit_model <- fitFPLS_DA(data = spectra100, method = "bsplines", nbasis = 1050)

# Inspect the optimal number of PLS components selected by cross-validation
fit_model$ncompOpt
```


---

### 3. Evaluation and Prediction (predict.FPLS_DA)

The prediction method maps new spectral matrices onto the functional basis and the PLS space defined during training, computing posterior group probabilities via the LDA component. The threshold argument allows users to flag ambiguous spectra under a "Doubty" classification tag.

```{r predict-model, message = FALSE, warning = FALSE}
# Prediction on the training data
predictions <- predict(object = fit_model, newdata = spectra100 )
# predictions <- predict.FPLS_DA(object = fit_model, newdata = spectra100 )

# Check the first few predicted classes
head(predictions$class)

# Checking accuracy
accuracy <- mean(predictions$class == rownames(spectra100))
print(paste("Accuracy : ", round(accuracy, 2)))

table(predictions$class , rownames(spectra100))
```

---

### 4. Simulating Functional Spectral Profiles (SimulateSpectrum)

To evaluate classifier stability under realistic conditions, SimulateSpectrum performs a Functional Principal Component Analysis (FPCA) selected via the AIC criterion. A K-means clustering step partitions the latent space, and a resampling engine generates $n$ new continuous spectra using the Karhunen-Loeve expansion combined with an adaptive noise structure. Physical non-negativity constraints are enforced via a pmax(y, 0) rectification step.

Three clustering methods are available via `clust_method`: K-means (default), hierarchical clustering (`"hclust"`), and Gaussian Mixture Models (`"gmm"`).

Setting `plot = TRUE` automatically outputs a $2 \times 2$ diagnostic dashboard.

```{r simulate-spectra, message = FALSE, warning = FALSE, fig.width  = 10, fig.height = 6, out.width  = "100%", fig.align  = "center"}

# Simulate 40 new spectra across 3 latent clusters with diagnostic plots enabled
simulated_spectra <- SimulateSpectrum(data = spectra100, n = 60, k = 3, 
                                      clust_method = "kmeans", factorNoise = 1, plot = TRUE)
```

---

### 5. Visualizing Discriminant Latent Spaces

The package includes advanced graphic functions to inspect microbial strain segregation within the projection subspaces.2D Score Projections with Class Probabilities (plotSpectra2D)This function displays the score map along the first two PLS dimensions. Enabling show_proba = TRUE applies strict discrete opacity tiers ($1.0, 0.6,$ or $0.2$) based on the maximum posterior LDA probability, helping users visually isolate taxonomic uncertainty zones.

```{r plot-2d, message = FALSE, warning = FALSE}

plotSpectra2D(fit_model, comp = c(1, 2), show_proba = TRUE , palette = NULL)
plotSpectra2D(fit_model, comp = c(1, 2), show_proba = FALSE, palette = NULL)
```


**Interactive 3D Scores Cloud (plotSpectra3D)**
To analyze complex cluster sub-structures, plotSpectra3D builds an interactive 3D scatter plot using plotly. Axes automatically integrate individual explained variance percentages, and plotting ranges expand by 10% to prevent marker edge truncation.

```{r plot-3d, message = FALSE, warning = FALSE}
# Generates an interactive 3D plot (supporting zoom and rotation)
plotSpectra3D(fit_model, comp = c(1, 2, 3), show_proba = FALSE, palette = NULL)
```

**UMAP Non-linear Embedding Projection (plotSpectraUMAP)**
When high biological proximity between distinct strains creates non-linear trajectories within the PLS score space, plotSpectraUMAP projects the complete PLS score matrix onto a 2D UMAP layout.

```{r plot-umap, message = FALSE, warning = FALSE}

plotSpectraUMAP(fit_model, n_neighbors = 10, min_dist = 0.01, show_proba = TRUE)
plotSpectraUMAP(fit_model, n_neighbors = 10, min_dist = 0.01, show_proba = FALSE)
```

**Model Diagnostics: Variance and Biomarker Trackers (plotPLSvar & plotPLSweights)**
To assist in model interpretation, plotPLSvar tracks individual variance decay across components, while plotPLSweights plots loading profiles across mass channels to expose the specific $m/z$ biomarkers driving the strain discrimination.

```{r plot-diagnostics, message = FALSE, warning = FALSE}
# Variance scree plot
plotPLSvar(fit_model)

# Identify influential m/z regions on Component 1
plotPLSweights(fit_model, comp = 1)
```

---

### 6. Model Summary

Once the model is fitted, `summary()` provides a structured diagnostic overview of the entire FPLS-DA pipeline in a single call. It covers four sections: the functional decomposition settings, the PLS dimension reduction statistics, the training data composition, and the in-sample classification performance.

```{r summary-model, message = FALSE, warning = FALSE}
s <- summary(fit_model)
```

The function can also print a formatted report directly to the console. The returned object `s` also allows programmatic access to every computed statistic without re-running the summary.

```{r summary-access, message = FALSE, warning = FALSE}
# Overall accuracy
cat(sprintf("Overall in-sample accuracy : %.1f%%\n", s$accuracy * 100))

# Per-group recall
print(s$recall)

# Full confusion matrix
print(s$confusion)
```

---

### 7. Deep Learning Classification with Functional Neural Networks (`FNN`)

For complex discrimination tasks, `maldiscrim` provides a Functional Neural Network (FNN) pipeline implemented in Python via `reticulate` and TensorFlow. The FNN captures non-linear spectral relationships that linear methods such as FPLS-DA may miss.

#### 7.1 Training the FNN 

Training a FNN model is computationally intensive. For demonstration purposes, this vignette loads a pre-trained model object directly from the package cache.

```{r load-fnn, message = FALSE, warning = FALSE}
data("fnn_model")
```

`fitFNN` trains the Functional Neural Network on a preprocessed spectral
matrix. The function handles Python environment setup, data formatting, model
compilation, and early stopping automatically.

```{r fnn-train, eval = FALSE}
# Full training example — not executed in this vignette (~1 hour)
# fnn_model_trained <- fitFNN(
#   data = spectra100,
#   trainSize = 0.8,
#   batchSize = 32L,
#   nbEpochs = 50L,
#   stepsPerEpoch = 16L,
#   patience = 5L,
#   monitor = "val_loss",
#   firstLayer = list(nFilters = 50L, nFunctions = 100L, resolution = 500L, basisType =
#     "Fourier", activation = "elu"),
#   secondLayer = list(nFilters = 5L, nFunctions = 70L, resolution = 200L, basisType =
#     "Fourier", activation = "elu"),
#   finalLayer = list(nFunctions = 50L, resolution = 19966L, basisType = "Fourier",
#     activation = "softmax", pooling = TRUE),
#   outputPath = "outputs/fnn",
#   verbose = TRUE
# )
```

#### 7.2 Network Architecture

Before training, `architectureFNN` allows the user to inspect and configure
the network layout — number of filters, kernel sizes, pooling sizes, and dense
units across the three main blocks (two functional convolutional blocks and a
final classification head). Calling it without arguments displays the default
configuration.

```{r fnn-architecture, eval = FALSE}
# Display and inspect the default FNN architecture
architectureFNN(fnn_model)
```
#### 7.3 Model Summary

`summary()` prints a structured report of the FNN configuration and in-sample
classification performance, following the same interface as `summary.FPLS_DA`.

```{r fnn-summary, message = FALSE, warning = FALSE}
s_fnn <- summary(fnn_model)
```

```{r fnn-summary-access, message = FALSE, warning = FALSE}
# Overall accuracy
cat(sprintf("FNN overall in-sample accuracy : %.1f%%\n", s_fnn$accuracy * 100))

# Per-group recall
print(s_fnn$recall)

# Full confusion matrix
print(s_fnn$confusion)
```


#### 7.4 Prediction 

`predict.FNN` projects new spectral data through the trained network and
returns the predicted class and the associated maximum posterior probability
for each sample. The optional `threshold` argument flags low-confidence
predictions as `"Doubty"`.

```{r fnn-predict, message = FALSE, warning = FALSE}
# Predict on training data (fitted values)
pred_fnn <- predict(fnn_model)

# First predicted classes
head(pred_fnn$class)

```

#### 7.5 Diagnostic Plots

**Training History (`plotTrainingFNN`)**

Monitors convergence and helps identify potential overfitting by displaying loss and accuracy curves for both the training and validation sets across epochs.

```{r fnn-history, message = FALSE, warning = FALSE}
plotTrainingFNN(fnn_model)
```

**In-Sample Confusion Matrix (`plotConfusionFNN`)**

Exposes cross-class ambiguities on the validation set as an annotated heatmap.

```{r fnn-confusion, message = FALSE, warning = FALSE}
plotConfusionFNN(fnn_model)
```

**Latent Space Projections (`plotUMAPFNN`, `plotMDSFNN`)**

Evaluate cluster separation across different activation levels (logits vs. softmax). Statistical confidence ellipses enclose distinct biological strains.

```{r fnn-umap, message = FALSE, warning = FALSE}
plotUMAPFNN(fnn_model, on = "logits",  seed = 42)
plotUMAPFNN(fnn_model, on = "softmax", seed = 42)
```

#### 7.6 Interpretability

**Interpretability: Grad-CAM Attribution (`plotGradCamFNN`)**

Highlights the spectral regions that most influenced the network's decision for a given sample and target class, using gradient-weighted class activation maps from the last functional convolutional layer.

```{r fnn-gradcam, message = FALSE, warning = FALSE}
# plotGradCamFNN(fnn_model, classIndex = 1, sampleIndex = 1)
```

**Interpretability: Occlusion Sensitivity (`plotOcclusionFNN`)**

Systematically masks spectral windows and monitors the resulting drop in class probability, mapping a sensitivity profile across the full m/z domain.

```{r fnn-occlusion, message = FALSE, warning = FALSE}
# plotOcclusionFNN(fnn_model, classIndex = 1, sampleIndex = 1, windowSize = 20)
```

**Interpretability: Filter Activations (`plotActivationFNN`)**

Visualizes the feature map outputs of each filter in the first functional convolutional layer for a chosen sample.

```{r fnn-activation, message = FALSE, warning = FALSE}
# plotActivationFNN(fnn_model, sampleIndex = 1)
```

**Interpretability: Activation Maximization (`plotActivationMaxFNN`)**

Uses gradient ascent to synthesize an idealized pseudo-spectrum that maximizes the prediction score for a specific target class, revealing what the network "expects" a given strain to look like.

```{r fnn-actmax, message = FALSE, warning = FALSE}
# plotActivationMaxFNN(fnn_model, classIndex = 1, steps = 50)
```
